---
title: "Bonanza Hares"
author: "Jacob Weverka"
date: "11/16/2020"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(janitor)
library(broom)
library(ggbeeswarm)
library(effsize)

```

# Introduction

Here, I will present exploratory analysis of data collected as part of a capture-recapture study  of snowshoe hares between 1998 and 2009 in the Tanana Valley in Alaska. Juvenile hares were collected from three sites within the study region, classified by sex, and several body measurements were taken. This analysis will focus only on data collected from juvenile hares. I will examine the number of juvenile hares caught through time, differences in body weights between male and female juvenile hares, and allometric relationships between hind foot size and body weight in juvenile hares. 



Read in data
```{r}
hares_data = read_csv(here::here("data/bonanza_hares.csv"))
```

# Data and Analyses Overview

The following analysis will focus on snowshoe hare data provided by [Knut et al (2017)](https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14). I will begin analysis by examining the number of juvenile snowshoe hares captured through time during the course of the study. I will then compare mean body sizes of male and female juvenile hares. I will then examine relationships between hind foot lenght and body weight in juvenile snowshoe hares.


Clean up data (dates)
```{r}

hares = hares_data %>%
  mutate(date = lubridate::mdy(date)) %>%
  mutate(year = lubridate::year(date))

juveniles = hares %>%
  filter(age == "j") %>%
  mutate(sex = replace_na(sex, "Unknown"))


```

# Exploratory findings


## Exploring counts of juvenile hares

Calculate number of juveniles and number of traps each year.

```{r}

jhares = hares %>%
  filter(age == "j") %>%
  group_by(year) %>%
  count(name = "juveniles")

traps = hares %>%
  group_by(year) %>%
  summarise(ntraps = n_distinct(trap))

juv_hares = jhares %>%
  full_join(traps, by = "year")


```

Make plot of juvenile captures by year

```{r fig.cap = "Juvenile hares caught in traps during each year of study"}
ggplot(juv_hares, aes(x = year, y = juveniles)) +
  geom_point(size = 2) +
  geom_line() +
  xlab("Year") +
  ylab("Number of Juvenile Hares Caught") +
  theme_bw() +
  scale_x_continuous(breaks = seq(1998, 2012, 2))
```

The mean number of juvenile hares caught annually was `r mean(juv_hares$n)` and the median was `r median(juv_hares$n)`. The most hares caught in a year was `r max(juv_hares$n)` and the least was `r min(juv_hares$n)`. Generally, the number of juveniles caught declined after the first two years of study. This could be indicative of a decline in the reproductive rate of the hare population, though it may not be appropriate to compare these counts across years, as different numbers of traps were placed each year, and many hare observations are missing age data.


## Vizualizing juvenile hare weights




Here, we're creating a dataframe of summary statistics for hare weights at each site and of each sex.

```{r}
juv_weight_sum = juveniles %>%
  group_by(sex, grid) %>%
  summarise(mean_weight = mean(weight, na.rm = TRUE), sd_weight = sd(weight, na.rm = TRUE)) %>%
  drop_na(sd_weight)


```
Now, we'll plot hare weights by sex and by site, with summary statistics overlayed.

```{r fig.cap= "Weights of juvenile hares by sex in each study site. Values shown in gray are individual hare weight. Mean weights are shown in brown, with one standard deviation represented by the error bar."}
grid_labs = c("Bonanza Black Spruce", "Bonanza Mature", "Bonanza Riparian")
names(grid_labs) = c("bonbs", "bonmat", "bonrip")

ggplot() +
  geom_beeswarm(data = juveniles, aes(x = sex, y = weight),
                size = 3,
                alpha = 0.3) +
  geom_point(data = juv_weight_sum,
             aes(x = sex, y = mean_weight),
             color = "chocolate",
             size = 5) +
  facet_wrap(.~grid,
             labeller = labeller(grid = grid_labs)) +
  geom_errorbar(data = juv_weight_sum,
                aes(x = sex,
                    ymin = mean_weight - sd_weight,
                    ymax = mean_weight + sd_weight),
                color = "chocolate") +
  scale_x_discrete(name = "Sex",
                   labels = c("f" = "Female", "m" = "Male", "Unknown" = "Not Recorded")) +
  ylab("Weight (grams)") +
  theme_bw()


```

From this graph, it appears that male juvenile hares had a higher mean body weight than females at each site. Still, it is not clear whether this is a meaningful or statistically significant difference, since the difference in means is less than a standard deviation for each site.




## Weight comparison between female and male hares



Here is a finalized table containing the mean, standard deviation, and sample size of juvenile male and female hares

```{r}

sex_summary = juveniles %>%
  group_by(sex) %>%
  summarise(mean_weight = mean(weight, na.rm = TRUE), sd_weight = sd(weight, na.rm = TRUE), n()) %>%
  mutate(sex = case_when(
    sex == "f" ~ "Female",
    sex == "m" ~ "Male",
    sex == "Unknown" ~ "Unknown"
  ))

knitr::kable(sex_summary, col.names = c("Sex", "Mean Weight", "Standard Deviation of Weight", "Sample Size"), caption = "Means and standard deviations of body weight for each sex of snowshoe hare")



```

Exploratory graphs to examine normality (graphs not shown, but code included)
```{r echo=TRUE, eval = FALSE }
ggplot(juveniles, aes(x = weight)) +
  geom_histogram() +
  facet_wrap(~sex)

ggplot(juveniles, aes(sample = weight)) +
  geom_qq() +
  facet_wrap(~sex)



```
The distribution of body weights does not appear exactly normal, based on the histograms and the qqplots. Still, the sample size is high enough that we should still be able to use parametric tests to look at differences between means.


Creating separate vectors for female and male weights

```{r}
juv_males = juveniles %>%
  filter(sex == "m") %>%
  pull(weight)

juv_females = juveniles %>%
  filter(sex == "f") %>%
  pull(weight)


```



A two sample t-test to compare means
```{r}
sex_t = t.test(juv_males, juv_females)
sex_diff_t = tidy(sex_t)
sex_t

```
Cohen's d test of effect size - difference in mean size of male and female snowshoe hares.
```{r}

sex_cohen = cohen.d(juv_males, juv_females, na.rm = TRUE)
sex_cohen

```
From our Welch two-sample t-test, we estimate the mean male body weight to be `r sex_diff_t$estimate1` and the mean female body weight to be `r sex_diff_t$estimate2`, for a difference of `r sex_diff_t$estimate1 - sex_diff_t$estimate2`. The p-value from this t test is `r sex_diff_t$p.value`, which mean that it is rather unlikely that male and female hares have the same mean body weight. Still, the estimated Cohen's d statistic is only `r sex_cohen$estimate`, which suggests that the difference in size between males and females is fairly small, even if it is statistically significant.


## Relationship between juvenile weight & hind foot length

Here, I'll start exploring the relationship between hind foot length and body weight by plotting these variables in a scatterplot.

```{r fig.cap= "Snowshoe hare hind foot lenght plotted against body weight"}
ggplot(juveniles, aes(x = hindft, y = weight)) +
  geom_point() +
  xlab("Hind Foot Length (mm)") +
  ylab("Weight (g)") +
  theme_bw()
```

Trying out a linear model relating hind foot length to body weight in juvenile snowshoe hares.
```{r}
hare_mod = lm(weight ~ hindft, data = juveniles)
summary(hare_mod)

hm = tidy(hare_mod)
hm2 = glance(hare_mod)

```
Now, examining correlation between hind foot length and weight for juvenile hares.
```{r}
hare_cor = cor.test(juveniles$hindft, juveniles$weight)
hare_cor
```
The plot and the linear model show a positive relationship between snowshoe hare hind foot length and body weight. From the linear model, a slope of `r hm$estimate[2]` indicates that for every additional 1 mm in hind foot length, a hare will average `r hm$estimate[2]` g more in body weight. With an r-squared value of `r hm2$r.squared`, variation in hind foot length explains about `r round(hm2$r.squared, 3) * 100` % of variation in body weight. With a Pearson's r correlation of `r hare_cor$estimate`, there is moderate correlation between these two variables. However, the diagnostic plots indicate some potential violations of linear model assumptions. In particular, the assumption of homoskedasticity might be broken here, as it appears that residuals are generally larger for larger values of hind foot length. 



Diagnostic plots for the linear model

```{r include = FALSE, echo=TRUE}
plot(hare_mod)
```

# Summary

In conclusion, the most juvenile hares were caught early in the study, and fewer were caught in subsequent years. Mean male body weight was larger than mean female body weight, but the difference in these means was small. Body weight appeared positively correlated to hind foot length in juvenile snowshoe hares, though it is not clear that a ordinary-least-squares linear model is appropriate for these data.

# Citations


Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14
































